---
title: "Variable contingency and affiliation prediction"
author: "Clemens Schmid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variable contingency and affiliation prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
devtools::load_all()
library(reshape2)
library(ggplot2)
library(corrplot)
library(dplyr)
library(igraph)
library(ca)
```

This Vignette shows a workflow for archaeologists workig on burial sites. Starting point is a data.frame with burials (objects) and their classified attributes (grave goods, orientation, size, sex etc.). An example dataset with 50 fictional burials and 18 stereotypical attributes is provided in `data/bs1`. Here's an extract of bs1:  

```{r, echo=FALSE}
bs1[1:10,9:16]
```

For bivariate analysis it's useful to use categorized data. In the context of burialsites the simple information wether a grave good is present or absent in a certain grave is often already sufficient. To reduce our dataset to this information the function `booleanize()` can be applied. Empty graves and attributes that never appeare can be removed with `delempty()`. 

```{r, echo=FALSE, echo=TRUE}
bs <- quantaar::booleanize(bs1)
bs <- quantaar::delempty(bs)
bs[1:10,9:16]
```

Let's look at the appearences of the individual grave goods with `presencecount()`. While most goods are frequent, *pottery_4* is quite rare - too rare to be meaningfully analyzed with chi square statistics. 

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=5}
presencematerial <- quantaar::presencecount(bs[,7:18], dim = 1)

presence.m <- reshape2::melt(presencematerial)
ggplot(presence.m,
       aes(x = variable,
           y = value)) +
  geom_bar(stat = "identity")
```

With `delrc()` we can delete every grave good that appeares less than ten times. As a consequence *pottery_4* is removed.

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=5}
bsred <- bs[,7:18]
bsred <- quantaar::delrc(bsred, climit = 10)

presencematerial <- quantaar::presencecount(bsred, dim = 1)
presence.m <- reshape2::melt(presencematerial)
ggplot(presence.m,
       aes(x = variable,
           y = value)) +
  geom_bar(stat = "identity")
```

Now we can calculate the bivariate relations of all attributes with the function `corrmat()`. `corrmat()` offers different correlation values. In this case we use the testdecision of the chi square test on a significance level of 2% and the phi coefficient. `rmnegcorr()` is used to remove significant but negative relationships.

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=7.5}
bsprep <- data.frame(bs[,1:6], bsred)
corrtablechi2test <- quantaar::corrmat(bsprep, method = "chi2", dim = 1, chi2limit = 0.02)
corrtablephi <- quantaar::corrmat(bsprep, method = "phi", dim = 1)
mastercorr <- quantaar::rmnegcorr(corrtablephi, bsprep, niv = 0.1, dim = 1)

col2 <- colorRampPalette(c("white","white", "chartreuse4"))
corrplot::corrplot(
  t(mastercorr),
  method = c("color")   
  )
```

`reltable()` creates a list of the significant relationships within the correlation table. We choose the phi coefficient to be the first correlation value and the chi square test decision to be the second. 

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=5, warning=FALSE}
signicorr <- quantaar::reltable(mastercorr, corrtablechi2test)
signicorr <- filter(signicorr, corrvalue2 == TRUE)

signicorr
```

The network of bivariate relationships can be plotted as a graph with the igraph-package. The male and the female clusters are clearly distinguishable. 

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=5, warning=FALSE}
signicorrmod <- data.frame(
  from = signicorr$namevar1, 
  to = signicorr$namevar2, 
  weight = signicorr$corrvalue)

graphbasis <- igraph::graph.data.frame(signicorrmod, directed = TRUE)
plot.igraph(graphbasis)
```

Within *signicorr* it's easy to search for the directly to an individual attribute linked attributes (level 1 relations).

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=7.5, warning=FALSE}
mvar <- c("sex_male", "sex_female")

mvar1male <- dplyr::filter(
  signicorr, 
  namevar1 == mvar[1] | 
    namevar2 == mvar[1]
)

mvar1male
```

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=7.5, warning=FALSE}
mvar1female <- dplyr::filter(
  signicorr, 
  namevar1 == mvar[2] | 
    namevar2 == mvar[2]
)


mvar1female
```

`predictvo()` uses the initial data.frame and *signicorr* to make a prediction about whether an object could contain a variable or not based on cross-references. Useful to determine for example the sex of buried individuals based on their grave goods when no anthropological determination is availbable.

```{r, echo=TRUE, message=FALSE, fig.width=7.5, fig.height=5, warning=FALSE}
predictgen <- quantaar::predictvo(bsprep, signicorr, mvar, level = 1)

sexprediction <- data.frame(
  m = predictgen[,1], 
  w = predictgen[,3],
  statsex = NA,
  mtat = bsprep$sex_male, 
  wtat = bsprep$sex_female,
  names = make.names(rownames(bsprep))
  )

for (i in 1:length(sexprediction[,1])){
  if (sexprediction$m[i] >= 1.5*sexprediction$w[i]){
    sexprediction$statsex[i] <- "m"
  } else if (sexprediction$w[i] >= 1.5*sexprediction$m[i]) {
    sexprediction$statsex[i] <- "w"
  } else {
    sexprediction$statsex[i] <- "uncertain"
  }
}

sexprediction
```